Supplementary Material

### 2025-05-23 manuscript/manuscript.qmd
manuscript update

---
title: "Shifting paradigms in Ocean Color: Bayesian Inference for Uncertainty-Aware Chlorophyll Estimation "
format:
    agu-pdf:
        keep-tex: true
    agu-html: default
author:  
  - name: Erdem M. Karaköylü
    orcid: 0000-0002-6156-1720
    corresponding: true
    email: erdemk@protonmail.com
    roles:
      - Investigator
      - Software
      - Visualization
      - Project administration
    affiliations:
      - Independant Researcher
  - name: Susanne E. Craig
    orcid: 0000-0002-8963-0951
    corresponding: false
    email: susanne.e.craig@nasa.gov
    roles:
      - Co-Investigator
    affiliations:
      - NASA/UMBC
abstract: Placeholder
plain-language-summary: Placeholder
keywords: ['Bayesian modeling', 'Ocean Color', ]
key-points:
  - Bayesian models have not been fully adopted in Ocean Color remote sensing yet.
  - These models are a principled way to take uncertainty into account
  - 
bibliography: references.bib
citation:
  container-title: Geophysical Research Letters
keep-tex: true
number-sections: true
date: last-modified
---

@@

This model adds a measurement-method-specific dispersion term to the structure of Model 6, capturing the known increase in variability associated with fluorescence-based chlorophyll estimates.

- Additional noise component applied when chl_flag = fluorescence  
- All other structure inherited from Model 6

@@

Model 6 - Similar to Model 5; multi-level hierarchical linear model partial pooling across $MBR$  groups (Rrs411, Rrs443, Rrs489, Rrs510) for both model mean and variance. However the linear variance model gets an additional parameter $γ$ that allows for the model dispersion to depend on the chlorophyll measurement method (fluorescence or HPLC). While this doesn't affect future data prediction, for which ground truth mearuement are optional, it does allow the development of insight on the effect of chloroophyll measurement method on model ucertainty.
:::

@@

*Model Fitting* - The model posterior was approximated using the No U-Turn Sampler (NUTS), an efficient and mature variant of Hamiltonian Monte Carlo (HMC), which uses particle physics to sample the posterior. Sampler settings were as follows. Four independent sampling runs referred to as *chains* were used to assess sampler convergence. Prior to sampling, 1000 tuning steps were used for each chain to maximize sampler efficiency. The 2000 samples were recorded for each chain totalling 8000 samples. I assessed sampler convergence using the Gelman-Rubin statistic, $\hat{R}$ and effective sample size($ESS$). Convergence is acceptable for $\hat{R}$ < 1.01 and $ESS > 1000$. Full trace plots and diagnostics are included in the Supplementary Material section. Goodness of fit was assessed using *Posterior plots*, which show  *Posterior Predictive Checks, which show the expected

>>>>

---
title: "Shifting paradigms in Ocean Color: Bayesian Inference for Uncertainty-Aware Chlorophyll Estimation "
format:
    agu-pdf:
        keep-tex: true
    agu-html: default
author:  
  - name: Erdem M. Karaköylü
    orcid: 0000-0002-6156-1720
    corresponding: true
    email: erdemk@protonmail.com
    roles:
      - Investigator
      - Software
      - Visualization
      - Project administration
    affiliations:
      - Independant Researcher
  - name: Susanne E. Craig
    orcid: 0000-0002-8963-0951
    corresponding: false
    email: susanne.e.craig@nasa.gov
    roles:
      - Co-Investigator
    affiliations:
      - NASA/UMBC
abstract: Placeholder
plain-language-summary: Placeholder
keywords: ['Bayesian modeling', 'Ocean Color', ]
key-points:
  - Bayesian models have not been fully adopted in Ocean Color remote sensing yet.
  - These models are a principled way to take uncertainty into account
  - 
bibliography: references.bib
citation:
  container-title: Remote Sensing of Environment
keep-tex: true
number-sections: true
date: last-modified
---

@@

This model adds a measurement-method-specific dispersion term to the structure of Model 6, capturing the known increase in variability associated with fluorescence-based chlorophyll estimates. A reasonable assumption is that HPLC measurements are less noisy than fluorescence. Thus $σ$ is given the expression:

@@

Model 6 - Similar to Model 5; multi-level hierarchical linear model partial pooling across $MBR$  groups (Rrs411, Rrs443, Rrs489, Rrs510) for both model mean and variance. However the linear variance model gets an additional parameter $γ$ that allows for the model dispersion to depend on the chlorophyll measurement method (fluorescence or HPLC). While this does not necessarily affect future , for which ground truth mearuement are optional, it does allow the development of insight on the effect of chloroophyll measurement method on model ucertainty.
:::

@@

*Model Fitting* - The model posterior was approximated using the No U-Turn Sampler (NUTS), an efficient and mature variant of Hamiltonian Monte Carlo (HMC), which uses particle physics to sample the posterior; see the Supplementary Materials section for more details. Sampler settings were as follows. Four independent sampling runs referred to as *chains* were used to assess sampler convergence. Prior to sampling, 1000 tuning steps were used for each chain to maximize sampler efficiency. The 2000 samples were recorded for each chain totalling 8000 samples. I assessed sampler convergence using the Gelman-Rubin statistic, $\hat{R}$ and effective sample size($ESS$). Convergence is acceptable for $\hat{R}$ < 1.01 and $ESS > 1000$. Full trace plots and diagnostics are included in the Supplementary Material section. Goodness of fit was assessed using *Posterior plots*, which show  *Posterior Predictive Checks, which show the expected

### 2025-05-23 manuscript/supplement.qmd
manuscript update

### Model Description and Priors
- brief word description of the idea behind the model and the priors used.

\begin{align*}
    \alpha_\mu &\sim \text{Normal}(0, 1) \\
    \alpha_\sigma &\sim \text{Exponential}(1) \\
    \beta_\mu &\sim \text{Normal}(0, 1) \\
    \beta_\sigma &\sim \text{Exponential}(1) \\
    \sigma_{\alpha_\mu} &\sim \text{Normal}(0, 1) \\
    \sigma_{\alpha_\sigma} &\sim \text{Exponential}(1) \\
    \sigma_{\beta_\mu} &\sim \text{Normal}(0, 1) \\
    \sigma_{\beta_\sigma} &\sim \text{Exponential}(1) \\
    \alpha_j &\sim \text{Normal}(\alpha_\mu, \alpha_\sigma) \quad \text{for } j = 1, \dots, J \\
    \beta_j &\sim \text{Normal}(\beta_\mu, \beta_\sigma) \quad \text{for } j = 1, \dots, J \\
    \sigma_{\alpha_j} &\sim \text{Normal}(\sigma_{\alpha_\mu}, \sigma_{\alpha_\sigma}) \quad \text{for } j = 1, \dots, J \\
    \sigma_{\beta_j} &\sim \text{Normal}(\sigma_{\beta_\mu}, \sigma_{\beta_\sigma}) \quad \text{for } j = 1, \dots, J \\
    \mu_i &= \alpha_{g_i} + \beta_{g_i} \cdot \log\text{MBR}_i \quad \text{for } i = 1, \dots, N \\
    \log \sigma_i &= \sigma_{\alpha_{g_i}} + \sigma_{\beta_{g_i}} \cdot \log\text{MBR}_i \\
    \sigma_i &= \exp(\log \sigma_i) \\
    \log\text{Chl}_i &\sim \text{TruncatedNormal}(\mu_i, \sigma_i, -3, 3.2)
\end{align*}

@@

```{figure} figs/ppc_example.png
---
width: 85%
fig-cap: "Figure S1: Posterior predictive check comparing observed and simulated chlorophyll-a distributions across groups."
---

>>>>

### Model Description and Priors
- brief word description of the idea behind the model and the priors used.
### Mathematical Description
\begin{align*}
    \alpha_\mu &\sim \text{Normal}(0, 1) \\
    \alpha_\sigma &\sim \text{Exponential}(1) \\
    \beta_\mu &\sim \text{Normal}(0, 1) \\
    \beta_\sigma &\sim \text{Exponential}(1) \\
    \sigma_{\alpha_\mu} &\sim \text{Normal}(0, 1) \\
    \sigma_{\alpha_\sigma} &\sim \text{Exponential}(1) \\
    \sigma_{\beta_\mu} &\sim \text{Normal}(0, 1) \\
    \sigma_{\beta_\sigma} &\sim \text{Exponential}(1) \\
    \alpha_j &\sim \text{Normal}(\alpha_\mu, \alpha_\sigma) \quad \text{for } j = 1, \dots, J \\
    \beta_j &\sim \text{Normal}(\beta_\mu, \beta_\sigma) \quad \text{for } j = 1, \dots, J \\
    \sigma_{\alpha_j} &\sim \text{Normal}(\sigma_{\alpha_\mu}, \sigma_{\alpha_\sigma}) \quad \text{for } j = 1, \dots, J \\
    \sigma_{\beta_j} &\sim \text{Normal}(\sigma_{\beta_\mu}, \sigma_{\beta_\sigma}) \quad \text{for } j = 1, \dots, J \\
    \mu_i &= \alpha_{g_i} + \beta_{g_i} \cdot \log\text{MBR}_i \quad \text{for } i = 1, \dots, N \\
    \log \sigma_i &= \sigma_{\alpha_{g_i}} + \sigma_{\beta_{g_i}} \cdot \log\text{MBR}_i \\
    \sigma_i &= \exp(\log \sigma_i) \\
    \log\text{Chl}_i &\sim \text{TruncatedNormal}(\mu_i, \sigma_i, -3, 3.2)
\end{align*}

## Model 6:

### Mathematical Description

\begin{align*}
    \alpha_\mu &\sim \text{Normal}(0, 1) \\
    \alpha_\sigma &\sim \text{Exponential}(1) \\
    \beta_\mu &\sim \text{Normal}(0, 1) \\
    \beta_\sigma &\sim \text{Exponential}(1) \\
    \sigma_{\alpha_\mu} &\sim \text{Normal}(0, 1) \\
    \sigma_{\alpha_\sigma} &\sim \text{Exponential}(1) \\
    \sigma_{\beta_\mu} &\sim \text{Normal}(0, 1) \\
    \sigma_{\beta_\sigma} &\sim \text{Exponential}(1) \\
    \alpha_j &\sim \text{Normal}(\alpha_\mu, \alpha_\sigma) \quad \text{for } j = 1, \dots, J \\
    \beta_j &\sim \text{Normal}(\beta_\mu, \beta_\sigma) \quad \text{for } j = 1, \dots, J \\
    \sigma_{\alpha_j} &\sim \text{Normal}(\sigma_{\alpha_\mu}, \sigma_{\alpha_\sigma}) \quad \text{for } j = 1, \dots, J \\
    \sigma_{\beta_j} &\sim \text{Normal}(\sigma_{\beta_\mu}, \sigma_{\beta_\sigma}) \quad \text{for } j = 1, \dots, J \\
    \gamma &\sim \text{Normal}(0, 1) \quad \text{(fluorescence-specific noise effect)} \\
    \mu_i &= \alpha_{g_i} + \beta_{g_i} \cdot \log\text{MBR}_i \quad \text{for } i = 1, \dots, N \\
    \log \sigma_i &= \sigma_{\alpha_{g_i}} + \sigma_{\beta_{g_i}} \cdot \log\text{MBR}_i + \gamma \cdot (1 - \text{ChlType}_i) \\
    \sigma_i &= \exp(\log \sigma_i) \\
    \log\text{Chl}_i &\sim \text{TruncatedNormal}(\mu_i, \sigma_i, -3, 3.2)
\end{align*}

## Model Fitting

### The Posterior Distribution

Mathematically the **posterior distribution** of the model parameters, given the observed data, is defined by Bayes' Theorem. For the simpler *Model 1* this looks like:

$$p(\alpha, \beta_1, \beta_2, \beta_3, \beta_4, \sigma | X) = \frac{p(X | \alpha, \beta_1, \beta_2, \beta_3, \beta_4, \sigma) \, p(\alpha, \beta_1, \beta_2, \beta_3, \beta_4, \sigma)}{p(X)}$$

Where:
* $p(\text{data} | \alpha, \beta_1, \beta_2, \beta_3, \beta_4, \sigma)$ is the **likelihood function**, which is the product of the individual likelihoods for each observed data point $y_i$:
    $$
    p(\text{data} | \alpha, \beta_1, \beta_2, \beta_3, \beta_4, \sigma) = \prod_{i=1}^{N} \text{TruncatedNormal}(y_i | \mu_i, \sigma, -3, 3.3)
    $$
    with
    $$
    \mu_i = \alpha + \beta_1 X_i + \beta_2 X_i^2 + \beta_3 X_i^3 + \beta_4 X_i^4
    $$
* $p(\alpha, \beta_1, \beta_2, \beta_3, \beta_4, \sigma)$ is the **joint prior distribution** of all parameters, which with the underlying assumption that they are conditionally independent, is the product of their individual prior distributions:
    $$
    p(\alpha, \beta_1, \beta_2, \beta_3, \beta_4, \sigma) = p(\alpha) \, p(\beta_1) \, p(\beta_2) \, p(\beta_3) \, p(\beta_4) \, p(\sigma)
    $$
    From your model:
    * $p(\alpha) = \text{Normal}(0, 1)$
    * $p(\beta_j) = \text{Normal}(0, 1)$ for $j=1,2,3,4$
    * $p(\sigma) = \text{Gamma}(2, 2)$
* $p(X)$ is the **marginal likelihood** (also known as the evidence), which is a normalizing constant:
    $$
    p(X) = \int p(X | \alpha, \beta_1, \beta_2, \beta_3, \beta_4, \sigma) \, p(\alpha, \beta_1, \beta_2, \beta_3, \beta_4, \sigma) \, d\alpha \, d\beta_1 \, d\beta_2 \, d\beta_3 \, d\beta_4 \, d\sigma
    $$

With many parameters, these integrals become quick intractable. Thus for often then none, the full posterior distribution will have to be approximated. The most accurate approximation is Markov Chain Monte Carlo (MCMC) sampling  based on accepted/rejected probabilistic jumps in the posterior space. MCMC algorithms have become quite since the early Metropolis-type algorithms. In this study I used the NUTS sampler, discussed next.

### Approximating the Posterior with HMC and the NUTS Sampler
Hamiltonian Monte Carlo (HMC) is an MCMC method that conceptualizes parameters as particles moving on a potential energy surface derived from the negative log-posterior. It introduces momentum variables and simulates Hamiltonian dynamics through "leapfrog steps" to propose new parameter values, allowing for more efficient exploration of complex parameter spaces compared to simpler MCMC techniques. The No-U-Turn Sampler (NUTS) enhances HMC by adaptively determining the optimal number of leapfrog steps in each iteration, preventing inefficient "U-turns" in the sampling trajectory. This makes NUTS a highly efficient and robust algorithm for approximating posterior distributions. The pre-print article by @betancourt2018

### Posterior Predictive

@@

```{figure} images/ppc_example.png
---
width: 85%
fig-cap: "Figure S1: Posterior predictive check comparing observed and simulated chlorophyll-a distributions across groups."
---
